No proof nets for MLL with units\\[5pt]\Large Proof equivalence in MLL is \textsc{pspace}-complete

### 2014-01-20 main/LICS.tex
Updated intro

%% STILL TO ADD:
%% - MLL proof equivalence is the word problem for *-autonomous categories
%% - IMLL proof equivalence is polynomial-time

An important element of the theory of linear logic \cite{Girard-1987} is the idea that proofs should be considered modulo the permutation of inferences.
%; that the chosen order of permutable inferences is inessential, or `bureaucracy'.
%contributes no essential proof content, and should be considered `bureaucracy'.
%
%The idea is supported by the categorical semantics of linear logic, in which proofs up to permutations denote the same morphism. 
%
\emph{Proof equivalence} is the problem of deciding whether two proofs are equal modulo permutations.

The proof equivalence problem is closely related to the notion of \emph{proof nets}.
%
These are canonical, geometric representations of proof that factor out permutations, so that proofs are equivalent if and only if they have the same underlying proof net.
%
Canonical proof nets have been found for several fragments of linear logic: the multipicative fragment without units \cite{Girard-1987,Danos-Regnier-1989}, the combined multiplicative-additive fragment without units \cite{Hughes-vanGlabbeek-2005}, and the additive fragment, including the additive units \cite{Heijltjes-2011}.
%
For the full multiplicative fragment, with units, several notions of proof net have been proposed \cite{BCST,Strassburger-lamarche-2004,HughesFreeStar}.
%
However, none is canonical, and whether a satisfactory notion of proof net can be found has remained an open question.

In this paper we establish that the proof equivalence problem for multiplicative linear logic with units is \textsc{pspace}-complete.
%
As a main corollary, this effectively rules out the possibility of finding a satisfactory notion of proof net for this fragment.

>>>>

\noindent The question of equivalence of proofs goes back to Lambek \cite{Lambek-1968}, who realised that the then-new tool of category-theoretic logic gave a notion of proof equivalence that was coarser and better-behaved than syntactic equality.

A striking technical innovation of linear logic was the introduction of proof nets \cite{Girard-1987,Danos-Regnier-1989}, which define a canonical form for proofs in the unitless fragment of \textsc{mll} -- two proofs are equivalent if and only if they have the same proof net -- so proof nets offer a simple decision procedure for proofs in this fragment. This naturally raises the question whether proof nets can be extended to work in the presence of units. There is a line of work in this direction, beginning with \cite{TrimbleThesis,BCST} and perhaps culminating with \cite{HughesFreeStar,HughesMLLProofNets}; but these proof nets are not canonical and must be identified up to a rewiring equivalence.

So the question remains whether there exist fully canonical proof nets for full \textsc{mll}. Canonical proof nets have been found for several other fragments of linear logic: the combined multiplicative-additive fragment without units \cite{Hughes-vanGlabbeek-2005}, and the additive fragment, including the additive units \cite{Heijltjes-2011}.

In this paper we establish that the proof equivalence problem for multiplicative linear logic with units is \textsc{pspace}-complete. It follows that there can be no satisfactory canonical notion of proof net.
